Active Code Completion

### 2012-03-20 icse12/graphite-icse12.tex
Cleaned up the references, added "palette reinvocation" in Section V, other small changes.

@@

In all of these systems, the code completion interface has remained primarily menu-based. When an item is selected, code is inserted immediately, without further input from the developer. These systems are also limited in that they are difficult to modularly extend. A fixed strategy determines the completions that are available, leaving library providers unable to specify new domain-specific or contextually-relevant items. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions. This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler. This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

@@

% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently. For each use case, we designed three mockup screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated. By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
\subsection{Survey Methods}
We recruit participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com\footnote{http://www.reddit.com/r/programming}. An additional 22 participants were recruited from the local computer science graduate student body.

@@

We first asked participants about their level of familiarity with several programming languages, on a five-point Likert scale ranging from ``None'' to ``Expert''. 61.1\% of the participants indicated that they were an expert in at least one language, and an additional 35.7\% were ``very familiar'' with at least one language. On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.

@@

Although our presentation of active code completion did not include any completely mouse-driven interfaces, several participants commented that the Color palette included interface elements taken from standard color dialog boxes that could only be manipulated using the mouse. These comments were generally severe in their condemnation of mouse-based interfaces in developer tools, consistent with our finding that a significant portion of our participants were familiar with editors like Vim that place a strong emphasis on keyboard shortcuts.

@@

The mockups we showed users were based on the Java language and the Eclipse development environment. As we showed, a number of participants preferred other languages or editors. Many of these participants made comments asking that the features we describe not be restricted to one specific editor (e.g. NetBeans users) and some also called for support for other programming languages (e.g. Ruby and C\#). Indeed, the palettes we demonstrated could be used with only slight modifications in a variety of programming languages, given suitable architectural support for porting palettes between editing environments.

@@

Our initial color palette was deemed overly complex by many participants (26), with many others (26) specifically saying that they would be satisfied with a simpler interface with colors listed next to their names. Other users wanted additional features, such as an ``eyedropper'' tool for selecting a color directly from an image on the screen. Similarly, the regular expression and SQL palettes were considered too simple by 12, and 15 participants respectively. User requests included syntax highlighting, a range of additional helpful tools for generating,  testing and sharing regular expressions, and the ability to browse databases more directly. Due to the wide range of user needs even within a single class, it may be that support for multiple palettes or a tabbed palette interface will be necessary in practice. To support incremental improvements to palettes based on user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.

@@

@@

\begin{itemize}
\item Because palettes are implemented in Javascript, any differences between Java and Javascript become evident. For example, Javascript color names are, in two cases, slightly different from the corresponding Java color name. The Javascript regular expression engine is also slightly different. Although a Java applet could be embedded to provide necessary interoperability in cases where these differences are critical, this is more difficult than it would be if the palettes were implemented using the same language as the code.
\item The palette user interface must stay within its bounding box -- user interface elements like pop-up menus (unless they are provided by the browser itself) are thus more difficult to implement and may require additional API support in the future.
\item Several use cases could benefit from greater access to the surrounding code, or even the surrounding project. Implementing this modularly is particularly difficult, and the Eclipse API does not easily allow for  serialization of the sum of its contextual knowledge for consumption by a palette.
\item Implementing a reinvocation mechanism that is robust to changes in the code and does not require the user to select the relevant information is difficult because this would require that the palette  parse all the code to find relevant information. This also would require richer insertion APIs that allow for  ``dead'' code to be removed. This puts a greater burden on palette developers and errors could affect user's code in subtle ways. The small inconvenience of having to select code before reinvoking a palette may be balanced by the knowledge that no other portions of the code will be modified inadvertently by that palette.
\end{itemize}

@@

We recruited 7 subjects from Computer Science Department and Institute for Software Research at Carnegie Mellon University. All the 7 subjects were PhD students. The subjects were randomly assigned into two groups: 4 subjects were assigned to the control group, and the other 3 subjects to the treatment group. There were six male participants and one female participant. Participants were compensated in the amount of 15 dollars for their participation.

@@

There were a total of 9 tasks to be completed in 45 minutes. The first 6 tasks involved writing regular expressions to validate various data formats (e.g. temperatures), and the remaining 3 tasks involved writing regular expressions to retrieve data from a document. The participants were allowed to move back and forth while doing the experiment, and they were allowed to use any external resources, including the internet, local console, and so forth (we did not observe any usage of programs other than the web browser, however). The only restriction we placed on their activity was that they were NOT allowed to directly search for the answer to a task online online. We omit descriptions of each individual question due to space limitations.

%
%\begin{itemize}
%	\item Data format validation tasks (total 6 tasks)
%	
%	\begin{enumerate}
%		\item US phone numbers
%		
%		For simplicity, it was assumed that any digit can be used in any place in the phone number, even though there is a notion of valid range for the area code.
%		
%		\item Times
%		
%		Hours and minutes only.
%		
%		\item Temperatures
%		
%		Only Fahrenheit and Celcius were used, and subjects were instructed to assume the degree symbol had not been used.
%		
%		\item Dates
%		
%		We did not place any restrictions on the format for dates. This task is quite difficult to solve generally. There are many valid formats for expressing dates, and the valid range of the day part depends on the specific year and month. Many participants were stuck on this task and chose to skip it and came back later.
%		
%		\item IPv4 addresses
%		
%		\item URLs starting with \texttt{http://} or \texttt{https://}
%		
%	\end{enumerate}
%	
%	\item Data searching tasks (total 3 tasks)
%	
%	\begin{enumerate}
%	\setcounter{enumi}{6}
%		\item Finding a fields declaration (member variable) in a Java class
%		
%		It was assumed that all the field declarations start with one of the access modifier keywords: \verb|public|, \verb|protected|, or \verb|private|.
%		
%		\item Finding a string literal in a Java file
%		
%		\item Finding an \texttt{img} tag in an HTML file
%		
%	\end{enumerate}
%	
%\end{itemize}
%
\subsubsection{Task Structure}

@@

\section{User Study Results}
After reading the prompt for the first task, all subjects began by searching for and reviewing documentation related to regular expressions. In all but one of these cases, the subject looked at the API documentation for the Java Pattern class during this initial review. The remaining subject, a member of the control group, referred to quick reference documentation provided by an external tool. In all cases, the documentation was left accessible in a browser window throughout the study, often displayed simultaneously alongside the coding window due to the large screen available for the study.

@@

\subsection{Visual Languages}
Because active code completion involves graphical user interface elements but ultimately generates textual source code representations, it can be considered a hybrid approach that borrows interaction techniques from visual languages while remaining compatible with conventional programming languages. This hybrid approach may help address some of the usability challenges previously associated with visual languages (cf. \cite{oai:CiteSeerXPSU:10.1.1.110.3146}).

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\bibliographystyle{IEEEtran}
%\bibliography{graphite-icse12}
\bibliography{../research}

>>>>

\usepackage{balance}

\clubpenalty = 10000
\widowpenalty = 10000
\displaywidowpenalty = 10000

@@

In all of these systems, the code completion interface has remained primarily menu-based. When an item is selected, code is inserted immediately, without further input from the developer. These systems are also limited in that they are difficult to modularly extend. A fixed strategy determines the completions that are available, leaving library providers unable to specify new domain-specific or contextually-relevant items. 
In this paper we propose a technique called {\it active code completion} that eliminates these restrictions \cite{ACC_VLHCC}. This  makes developing and integrating a broad array of highly-specialized developer tools directly into the editor, via the familiar code completion command, significantly simpler. This technique is motivated by the evidence discussed above and further evidence provided in this paper that developers prefer, and make more effective use of, tools that do not require leaving the immediate editing environment.

@@

% In order to convey the active code completion concept clearly, we came up with three different use cases where we belive the active code completion technique will benefit developers frequently. For each use case, we designed three mockup screenshots and showed them in sequence: before invoking the palette, during the interaction with the palette, and after the palette is closed and code is generated. By doing this, the respondents could understand the concept precisely and give useful feedback relevant to the above questions.
\subsection{Survey Methods}
We recruit participants for this survey\footnote{https://www.surveymonkey.com/s/2GLZP8V} primarily from a popular programming-related discussion forum hosted on the popular website reddit.com \cite{reddit_programming}. An additional 22 participants were recruited from the local computer science graduate student body.

@@

We first asked participants about their level of familiarity with several programming languages, on a five-point Likert scale\footnote{``None'', ``Somewhat familiar'', ``Familiar'', ``Very familiar'', ``Expert''}. 61.1\% of the participants indicated that they were an expert in at least one language, and an additional 35.7\% were ``very familiar'' with at least one language. On average, participants rated themselves as very familiar with Java, C, C++ and JavaScript, familiar with C\#, Python and PHP and somewhat familiar with Visual Basic and Perl.

@@

Although our mockup screenshots did not include any completely mouse-driven interfaces, several participants commented that the Color palette included interface elements taken from standard color dialog boxes that could only be manipulated using the mouse. These comments were generally severe in their condemnation of mouse-based interfaces in developer tools, consistent with our finding that a significant portion of our participants were using editors like Vim that place a strong emphasis on keyboard shortcuts.

@@

The mockups we showed users were based on the Java and the Eclipse IDE. As we showed, a number of participants preferred other languages or editors. Many of these participants made comments asking that the features we describe be IDE and programming language independent. Indeed, the palettes we demonstrated could be used with only slight modifications in a variety of programming languages, given suitable architectural support for porting palettes between editing environments.

@@

Our initial color palette was deemed overly complex by many participants (26), with many others (26) specifically wanting a simpler interface with colors listed next to their names. Other users wanted additional features, such as an ``eyedropper'' tool for selecting a color directly from an image on the screen. Similarly, the regular expression and SQL palettes were considered too simple by 12, and 15 participants respectively. User requests included syntax highlighting, a range of additional helpful tools for generating, testing and sharing regular expressions, and the ability to browse databases more directly. Due to the wide range of user needs even within a single class, it may be that support for multiple palettes or a tabbed palette interface will be necessary in practice. To support incremental improvements based on user feedback, an architecture that makes deploying new and updated palettes relatively painless would also be valuable.

@@

\subsection{Palette Reinvocation}
To make further changes on the code generated by a Graphite palette, users can select the generated code and reinvoke the palette. The individual palette is responsible for parsing the selected code and filling the parameters back in the palette appropriately. When the user is satisfied with the new parameters and hits {\tt ENTER}, the selected code is replaced with a newly generated code.

@@

\begin{itemize}
\item Because palettes are implemented in Javascript, any differences between Java and Javascript become evident. For example, the color names are slightly different between Java and Javascript, and so are their regular expression engines. Although a Java applet could be used in cases where these differences are critical, this is still more difficult than it would be if the palettes were implemented using the same language as the code.
\item The palette user interface must stay within its bounding box -- user interface elements like pop-up menus (unless they are provided by the browser itself) are thus more difficult to implement and may require additional API support in the future.
\item Several use cases could benefit from greater access to the surrounding code, or even the surrounding project. Implementing this modularly is particularly difficult, and the Eclipse API does not easily allow for  serialization of the sum of its contextual knowledge for consumption by a palette.
\item Implementing a robust reinvocation mechanism that does not require the user to select the generated code is difficult, because this would require the palette to parse all the code to find relevant information, and also remove the ``dead'' code. This puts a greater burden on palette developers and errors could affect user's code in subtle ways. The small inconvenience of having to select code before reinvoking a palette may be balanced by the knowledge that no other portions of the code will be modified inadvertently by that palette.
\end{itemize}

@@

@@

There were a total of 9 tasks to be completed in 45 minutes. The first 6 tasks involved writing regular expressions to validate various data formats (e.g. temperatures), and the remaining 3 tasks involved writing regular expressions to retrieve data from a document. The participants were allowed to move back and forth while doing the experiment, and they were allowed to use any external resources, including the internet, local console, and so forth (we did not observe any usage of programs other than the web browser, however). The only restriction we placed on their activity was that they were NOT allowed to directly search for the answer to a task online. We omit descriptions of each individual question due to space limitations.

%
%\begin{itemize}
%	\item Data format validation tasks (total 6 tasks)
%	
%	\begin{enumerate}
%		\item US phone numbers
%		
%		For simplicity, it was assumed that any digit can be used in any place in the phone number, even though there is a notion of valid range for the area code.
%		
%		\item Times
%		
%		Hours and minutes only.
%		
%		\item Temperatures
%		
%		Only Fahrenheit and Celcius were used, and subjects were instructed to assume the degree symbol had not been used.
%		
%		\item Dates
%		
%		We did not place any restrictions on the format for dates. This task is quite difficult to solve generally. There are many valid formats for expressing dates, and the valid range of the day part depends on the specific year and month. Many participants were stuck on this task and chose to skip it and came back later.
%		
%		\item IPv4 addresses
%		
%		\item URLs starting with \texttt{http://} or \texttt{https://}
%		
%	\end{enumerate}
%	
%	\item Data searching tasks (total 3 tasks)
%	
%	\begin{enumerate}
%	\setcounter{enumi}{6}
%		\item Finding a fields declaration (member variable) in a Java class
%		
%		It was assumed that all the field declarations start with one of the access modifier keywords: \verb|public|, \verb|protected|, or \verb|private|.
%		
%		\item Finding a string literal in a Java file
%		
%		\item Finding an \texttt{img} tag in an HTML file
%		
%	\end{enumerate}
%	
%\end{itemize}
%
%\subsubsection{Task Structure}

@@

\section{Pilot Study Results}

As a pilot study, we recruited 7 PhD students from School of Computer Science at Carnegie Mellon University. The subjects were randomly assigned into two groups: 4 subjects were assigned to the control group, and the other 3 subjects to the treatment group. There were six male participants and one female participant. Participants were compensated in the amount of 15 dollars for their participation.

All participants had prior experience with both Java and regular expressions, but they were all somewhat rusty on the details of the API. After reading the prompt for the first task, all subjects began by searching for and reviewing documentation related to regular expressions. In all but one of these cases, the subject looked at the API documentation for the Java Pattern class during this initial review. The remaining subject, a member of the control group, referred to quick reference documentation provided by an external tool. In all cases, the documentation was left accessible in a browser window throughout the study, often displayed simultaneously alongside the coding window due to the large screen available for the study.

@@

\subsection{Visual Languages}
Because active code completion involves graphical user interface elements but ultimately generates textual source code representations, it can be considered a hybrid approach that borrows interaction techniques from visual languages while remaining compatible with conventional programming languages. This hybrid approach may help address some of the usability challenges previously associated with visual languages (cf. \cite{doi:10.1080/1049482940040202}).

@@

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)
\bibliographystyle{IEEEtran}
%\bibliography{graphite-icse12}
\balance \bibliography{../research}
